---
title: "Confounder and richness analysis"
author: "Christian Diener <mail (a) cdiener.com>"
output: 
  html_notebook: default
---

## Additional analyses

```{r}
library(mbtools)

original <- fread("../data/tests_genus.csv")
original$strategy <- "without"
original$response <- "single"
filtered <- fread("../data/tests_confounders.csv")
filtered$strategy <- "with"

vars <- c("diabetes_status", "auc_glucose", "num_risk_factors", "bmi", "percent_body_fat", "systolic_pressure", "diastolic_pressure")
genera <- c("Escherichia/Shigella", "Veillonella", "Anaerostipes", "Blautia")
tests <- rbind(original[variable %in% vars], filtered[variable %in% vars])
tests <- tests[genus %in% genera]
tests[, variable := factor(variable, levels=vars)]
tests[, strategy := factor(strategy, levels=c("without", "with"))]
```

### Confounding of associations:

Shown are the log2 fold changes when only considering the clinical variable without additional confounders (only gender as always used) and when using
additional confounding variables for the other groups. Basically if you see a horizontal line it would indicate no confounding effect. If the line drops or 
raises towards zero it indicates that the association is confounded. Error bars denotes the standard error of the log2 fold change (regression coefficient).

```{r, fig.width=8, fig.height=4}
library(ggplot2)
theme_set(theme_bw())

pl <- ggplot(tests, aes(x=strategy, y=log2FoldChange, col=genus, group=genus)) +
  geom_hline(yintercept=0, lty="dashed") +
  geom_linerange(aes(ymin=log2FoldChange - lfcSE, ymax=log2FoldChange + lfcSE)) +
  geom_point() + geom_line(alpha=0.5) + facet_wrap(~ variable, nrow=2) + 
  labs(x="confounding", y="log2 fold change", color="")
ggsave("confounding.svg", width=8, height=4)
pl
```

Seems that most associations are actually due to the disease state.

## Richness

Richness was calculated using the Chao1 index.

Just to see that association with richness does not depend on overall abundance...

```{r}
pl <- ggplot(original[variable == "richness"], aes(x=baseMean, y=log2FoldChange, col=(padj > 0.05) | is.na(padj))) +
  geom_hline(yintercept=0, lty="dashed") + geom_point() + scale_x_log10(breaks=c(10, 100, 1000, 10000)) + 
  labs(x="abundance [norm. reads]", y="log2 fold change") +
  guides(color=F)
ggsave("abundance_vs_richness_lfc.svg", width=5, height=5)
pl
```

Genera which are associated significantly with richness. Error bars denotes the standard error of the log2 fold change (regression coefficient).

```{r, fig.width=6, fig.height=5}
rich <- original[padj < 0.05 & variable == "richness"][order(log2FoldChange)]
rich[, genus := factor(genus, levels=unique(genus))]
pl <- ggplot(rich, aes(y=genus, x=log2FoldChange)) +
  geom_vline(xintercept=0, lty="dashed") +
  geom_errorbarh(aes(xmin=log2FoldChange - lfcSE, xmax=log2FoldChange + lfcSE), height=0.2) +
  geom_point(aes(col=baseMean), size=2) + 
  scale_fill_gradient(name="abundance", trans="log") + 
  labs(x="log2 fold change", y="", color="abundance")
ggsave("richness_genera.svg", width=6, height=5)
pl
```

We find Escherichia and Veillonella but not Anaerostipes and Blautia. So whatever the causality between Anaerostipes/Blautia and diabetes is it is not
simply a restoration of diversity.
